---
title: "Summary Paper: Retail Strategy and Analytics"
author: "Navneeth Vittal H R"
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---
```{r setup, include=FALSE} 
# set options for R markdown knitting 
knitr::opts_chunk$set(echo = TRUE) 
knitr::opts_chunk$set(linewidth=80) 
```

```{r knitr line wrap setup, include=FALSE} 
# set up line wrapping in MD knit output 
library(knitr) 
hook_output = knit_hooks$get("output") 
knit_hooks$set(output = function(x, options) 
  {    
    # this hook is used only when the linewidth option is not NULL
    if (!is.null(n <- options$linewidth))
      {        
        x = knitr:::split_lines(x)    
        # any lines wider than n should be wrapped
        if (any(nchar(x) > n))
          x = strwrap(x, width = n)
        x = paste(x, collapse = "\n")
      }    
  hook_output(x, options)
  }) 
```

## Load required libraries 

```{r 0 Load Libraries, results = 'hide'}

install.packages("ggplot2", lib = "~/R/win-library/4.2")
install.packages("ggmosaic", lib = "~/R/win-library/4.2")
install.packages("data.table", lib = "~/R/win-library/4.2")
install.packages("readr", lib = "~/R/win-library/4.2")
install.packages("readxl", lib = "~/R/win-library/4.2")
install.packages("tidyr", lib = "~/R/win-library/4.2")

#### Load required libraries 
library(ggplot2)
library(data.table)
library(ggmosaic)
library(readr) 
library(readxl)
library(tidyr)
```

## Chapter 1: Introduction

### Background

Understand the types of customers who purchase Chips and their purchasing behavior within the region and strategic plan for the chip category in the next half year.

```{r}
transactionData <- read_excel("~/Quantium/Dataset/QVI_transaction_data.xlsx")
customerData <- read_csv("~/Quantium/Dataset/QVI_purchase_behaviour.csv")
```

## Exploratory data analysis

### Examining transaction data

```{r Examine structure of transaction data}
trans_rows <- nrow(transactionData)  
trans_cols <- ncol(transactionData)

# Print the structure of transactionData
cat("Structure of transactionData:\n")
str(transactionData)
cat("\n")

```
Transaction dataset contains `r trans_rows` observations of `r trans_cols` variables. We can see that the date column in Transaction dataset is in an integer format. Let's change this to a date format.

```{r Convert DATE to date format} 
transactionData$DATE <- as.Date(transactionData$DATE, origin = "1899-12-30") 
```

```{r Summary of PROD_NAME}

unique_prods <- unique(transactionData$PROD_NAME)
n_unique_prods <- length(unique_prods)

# Count occurrences of each product
prod_counts <- table(transactionData$PROD_NAME)

# Display top 10 most frequent product
top_n <- 10
top_prods <- head(names(prod_counts[order(-prod_counts)]), top_n)
top_prods
```
Number of unique products are `r n_unique_prods` and top `r top_n` most frequent products are `r prod_counts[top_prods]` 

```{r Further examine PROD_NAME} 

# Further examine PROD_NAME
productWords <- unlist(strsplit(unique(transactionData$PROD_NAME), " "))

# Remove digits and special characters
cleaned_words <- productWords[!grepl("[0-9&]", productWords)]

# Count occurrences of each cleaned word
word_counts <- table(cleaned_words)

# Sort the distinct words by frequency of occurrence (highest to lowest)
sorted_words <- names(sort(word_counts, decreasing = TRUE))

# Display the sorted words
cat("Top most common words:\n")
print(sorted_words)

```


```{r }
#### Remove salsa products
transactionData$SALSA <- grepl("salsa", tolower(transactionData$PROD_NAME))
transactionData <- subset(transactionData, SALSA == FALSE)
transactionData$SALSA <- NULL 
```

```{r Inital summary}

# Use summary() to check summary statistics and nulls

numerical_cols <- sapply(transactionData, is.numeric)
numerical_data <- transactionData[, numerical_cols]

# Display summary statistics for numerical columns
summary_stats <- summary(numerical_data)
print("Summary statistics for numerical columns:")
print(summary_stats)

```
There are no nulls in the columns but product quantity appears to have an outlier
which we should investigate further. Let's investigate further the case where 200
packets of chips are bought in one transaction.

```{r }
# Identify outliers using a boxplot for each numerical column
par(mfrow = c(2, 3))
for (col in names(numerical_data)) {
  boxplot(numerical_data[[col]], main = col)
}
```
```{r}
# Filter the dataset to find the transactions where 200 packets of chips are bought
outlier_transactions <- subset(transactionData, PROD_QTY == 200)

# Display the transactions
print("Transactions where 200 packets of chips are bought:")
print(outlier_transactions)
```
There are two transactions where 200 packets of chips are bought in one transaction
and both of these transactions were by the same customer.

```{r }
# Extract the loyalty card number from the outlier transactions
customer_loyalty_card <- unique(outlier_transactions$LYLTY_CARD_NBR)

# Filter the dataset to see other transactions by this customer
other_transactions_by_customer <- subset(transactionData, LYLTY_CARD_NBR == customer_loyalty_card)

# Display other transactions by the customer
print("Other transactions by the customer:")
print(other_transactions_by_customer)
```

It looks like this customer has only had the two transactions over the year and is
not an ordinary retail customer. The customer might be buying chips for commercial
purposes instead. We'll remove this loyalty card number from further analysis.

```{r }
# Filter out the customer based on the loyalty card number
filtered_transaction_data <- subset(transactionData, !(LYLTY_CARD_NBR %in% customer_loyalty_card))

# Re-examine transaction data
print("Transaction data after filtering out the customer:")
print(filtered_transaction_data)
```

Now, let's look at the number of transaction lines over time to see
if there are any obvious data issues such as missing data.

```{r }
# Count the number of transactions by date
transaction_count_by_date <- table(transactionData$DATE)

# Create a summary of transaction count by date
transaction_count_summary <- data.frame(Date = as.Date(names(transaction_count_by_date)),
                                       Transaction_Count = as.numeric(transaction_count_by_date))

# Print the summary
print("Summary of transaction count by date:")
print(transaction_count_summary)
```
There's only 364 rows, meaning only 364 dates which indicates a missing date. Let's
create a sequence of dates from 1 Jul 2018 to 30 Jun 2019 and use this to create a
chart of number of transactions over time to find the missing date.

```{r fig.align = "center"}
# Create a summary of transaction count by date
transactions_by_day <- as.data.frame(table(transactionData$DATE))

# Rename the columns for clarity
colnames(transactions_by_day) <- c("Date", "Transaction_Count")

# Convert Date column to Date type
transactions_by_day$Date <- as.Date(transactions_by_day$Date)

# Plot the transaction count by date
library(ggplot2)
ggplot(transactions_by_day, aes(x = Date, y = Transaction_Count)) +
  geom_line() +
  labs(title = "Transaction Count by Date",
       x = "Date",
       y = "Transaction Count") +
  theme_minimal()

```
We can see that there is an increase in purchases in December and a break in late
December. Let's zoom in on this.

```{r fig.align = "center"}
# Load the required library
library(tidyr)

# Create a data frame with all dates from 1 Jul 2018 to 30 Jun 2019
all_dates_df <- data.frame(DATE = seq(as.Date("2018-07-01"), as.Date("2019-06-30"), by = "day"))

# Perform left join and replace N column with zeros
transactions_filled <- left_join(all_dates_df, transactions_by_day, by = c("DATE" = "Date")) %>%
  mutate(Transaction_Count = ifelse(is.na(Transaction_Count), 0, Transaction_Count))

# Plot transactions over time
ggplot(transactions_filled, aes(x = DATE, y = Transaction_Count)) +
  geom_line() +
  labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
  scale_x_date(breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

```
We can see that the increase in sales occurs in the lead-up to Christmas and that
there are zero sales on Christmas day itself. This is due to shops being closed on
Christmas day.

Now that we are satisfied that the data no longer has outliers, we can move on to
creating other features such as brand of chips or pack size from PROD_NAME. We will
start with pack size.

```{r Create pack size}
# Convert transactionData to a data.table
setDT(transactionData)

# Add PACK_SIZE column
transactionData[, PACK_SIZE := parse_number(PROD_NAME)]

# Check the count of each pack size
transactionData[, .N, PACK_SIZE][order(PACK_SIZE)]
```

The largest size is 380g and the smallest size is 70g - seems sensible!

```{r }
# Plot histogram of PACK_SIZE
ggplot(transactionData, aes(x = PACK_SIZE)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Pack Sizes",
       x = "Pack Size",
       y = "Number of Transactions") +
  theme_minimal()
```
Pack sizes created look reasonable.

```{r }
# Extract brand name from PROD_NAME column
transactionData <- transactionData %>%
  mutate(Brand = sapply(strsplit(PROD_NAME, " "), `[`, 1))

head(transactionData)
```

```{r }
transactionData[Brand == "Red", Brand := "RRD"]

transactionData[Brand == "Snbts", Brand := "Sunbites"]
transactionData[Brand == "Dorito", Brand := "Doritos"]

# Check the unique values in the Brand column to see if the adjustments were applied correctly
unique(transactionData$Brand)
```

### Examining customer data

Now that we are happy with the transaction dataset, let's have a look at the
customer dataset.

```{r 1 Exploratory data analysis}

custmon_rows <- nrow(customerData)
customer_cols <- ncol(customerData)

# Print the structure of customerData

cat("Structure of customerData:\n")
str(customerData)
```

Customer dataset contains `r custmon_rows` observations of `r customer_cols` variables.

```{r }
# Merge transaction data to customer data
data <- merge(transactionData, customerData, all.x = TRUE)
```

```{r Check for missing customer details}
null_counts <- colSums(is.na(data))
print("Null counts:")
print(null_counts)
```
Great, there are no nulls! So all our customers in the transaction data has been
accounted for in the customer dataset.

```{r Code to save dataset as a csv}
filePath <- "~/Quantium/Dataset/"
fwrite(data, paste0(filePath,"QVI_data.csv"))
```

## Data analysis on customer segments
Now that the data is ready for analysis, we can define some metrics of interest to
the client:
- Who spends the most on chips (total sales), describing customers by lifestage and
how premium their general purchasing behaviour is
- How many customers are in each segment
- How many chips are bought per customer by segment
- What's the average chip price by customer segment

```{r fig.width = 10, fig.align = "center"}
# Calculate total sales by LIFESTAGE and PREMIUM_CUSTOMER
sales_summary <- data %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  summarise(total_sales = sum(TOT_SALES))

# Plot total sales by LIFESTAGE and PREMIUM_CUSTOMER
ggplot(sales_summary, aes(x = LIFESTAGE, y = total_sales, fill = PREMIUM_CUSTOMER)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Lifestage", y = "Total Sales", title = "Total Sales by Lifestage and Premium Customer") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Sales are coming mainly from Budget - older families, Mainstream - young
singles/couples, and Mainstream - retirees

Let's see if the higher sales are due to there being more customers who buy chips.

```{r fig.width = 10, fig.align = "center"}
customer_count <- data %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  summarise(num_customers = n_distinct(LYLTY_CARD_NBR))

# Plot the number of customers by LIFESTAGE and PREMIUM_CUSTOMER
ggplot(customer_count, aes(x = LIFESTAGE, y = num_customers, fill = PREMIUM_CUSTOMER)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Lifestage", y = "Number of Customers", title = "Number of Customers by Lifestage and Premium Customer") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
There are more Mainstream - young singles/couples and Mainstream - retirees who buy
chips. This contributes to there being more sales to these customer segments but
this is not a major driver for the Budget - Older families segment.

Higher sales may also be driven by more units of chips being bought per customer.
Let's have a look at this next.

```{r fig.width = 10, fig.align = "center"}
units_per_customer <- data %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  summarise(avg_units_per_customer = mean(PROD_QTY))

# Plot the average number of units per customer by LIFESTAGE and PREMIUM_CUSTOMER
ggplot(units_per_customer, aes(x = LIFESTAGE, y = avg_units_per_customer, fill = PREMIUM_CUSTOMER)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Lifestage", y = "Average Units per Customer", title = "Average Units per Customer by Lifestage and Premium Customer") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Older families and young families in general buy more chips per customer

```{r fig.width = 10, fig.align = "center"}
avg_price_per_unit <- data %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  summarise(avg_price_per_unit = mean(TOT_SALES / PROD_QTY))

# Plot the average price per unit by LIFESTAGE and PREMIUM_CUSTOMER
ggplot(avg_price_per_unit, aes(x = LIFESTAGE, y = avg_price_per_unit, fill = PREMIUM_CUSTOMER)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Lifestage", y = "Average Price per Unit", title = "Average Price per Unit by Lifestage and Premium Customer") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Mainstream midage and young singles and couples are more willing to pay more per
packet of chips compared to their budget and premium counterparts. This may be due
to premium shoppers being more likely to buy healthy snacks and when they buy
chips, this is mainly for entertainment purposes rather than their own consumption.
This is also supported by there being fewer premium midage and young singles and
couples buying chips compared to their mainstream counterparts.

As the difference in average price per unit isn't large, we can check if this difference is statistically different.

```{r}
mainstream <- data %>%
  filter(PREMIUM_CUSTOMER == "Mainstream", LIFESTAGE %in% c("MIDAGE SINGLES/COUPLES", "YOUNG SINGLES/COUPLES"))

premium <- data %>%
  filter(PREMIUM_CUSTOMER == "Premium", LIFESTAGE %in% c("MIDAGE SINGLES/COUPLES", "YOUNG SINGLES/COUPLES"))

budget <- data %>%
  filter(PREMIUM_CUSTOMER == "Budget", LIFESTAGE %in% c("MIDAGE SINGLES/COUPLES", "YOUNG SINGLES/COUPLES"))

# Perform independent t-tests
t_test_mainstream_premium <- t.test(mainstream$TOT_SALES, premium$TOT_SALES)
t_test_mainstream_budget <- t.test(mainstream$TOT_SALES, budget$TOT_SALES)

# Print the results
print(t_test_mainstream_premium)
print(t_test_mainstream_budget)
```

The t-test results in a p-value of 2.2e-16, i.e. the unit price for mainstream,
young and mid-age singles and couples ARE significantly higher than
that of budget or premium, young and midage singles and couples.
